---
title: "ST 558 Final Project - EDA"
author: Chris Grace
format: 
  html:
    toc: true
    theme: cosmo
---

# Introduction

We'll be investigating a dataset containing measurements for bodies of water with the ultimate goal of predicting the potability (safeness to drink) of the water.

## Reading in the Data

We use the `read.csv` function to bring in our data:
```{r}
water <- read.csv("water_potability.csv")

head(water)
```

Let's check how many rows and columns our data set has:
```{r}
dim(water)
```
There are `r nrow(water)` rows in the data set, each with `r ncol(water)` columns. The preview of the table shows some missing values, so we'll use `colsums()` to examine this missingness:

```{r}
colSums(is.na(water))
```
There are a considerable number of missing values for `sulfate`, `ph`, and `trihalomethanes`. We could drop these predictors or remove the rows with missing observations.

Next, we'll go over each of the variables in the data set.

## Variables in the Dataset

The dataset has the following variables:

- `ph` : The pH of water (0 to 14)
- `Hardness`: The hardness of the water (mg/L)
- `Solids`: Total dissolved solids (ppm)
- `Chloramines`: Amount of chloramines (ppm)
- `Sulfate`: Amount of sulfate (ppm)
- `Conductivity`: Electrical conductivity ($\mu$S/cm)
- `Organic_carbon`: Amount of organic carbon (ppm)
- `Trihalomethanes`: Amount of trihalomethanes ($\mu$g/L)
- `Turbidity`: Haziness (NTU)
- `Potability`: Indicator for safety of human consumption (1 = safe, 0 = not safe)

Our ultimate goal is to **predict potability** given the values of the other variables. In the remainder of the exploratory data analysis, our goal is to explore the relationships in the data and determine which predictor variables we should use for our models.

# Numeric Summaries

Our first step is to compare the average value of each predictor when water is potable/non-potable. This should make any strong relationships abundantly clear and give us a good starting point.

I'll clean up the column names a bit before we compute the summary measures.
```{r}
suppressPackageStartupMessages(library(tidyverse))

# make names lowercase
names(water) <- tolower(names(water))

# rename vars i can't spell
water <-
  water |>
  rename("potable" = "potability",
         "trihalo" = "trihalomethanes") |>
  mutate(potable = factor(potable,
                          labels = c("non_potable", "potable")))
  

# compute summary table
num_sum <-
  water |>
  drop_na() |>
  group_by(potable) |>
  summarize(across(where(is.numeric), mean)) |>
  pivot_longer(cols = !potable) |>
  pivot_wider(names_from = potable,
              values_from = value)

# create difference column
num_sum <-
  num_sum |>
  mutate(pct_difference = abs(potable - non_potable)/((potable+non_potable) / 2)) |>
  arrange(desc(pct_difference))

num_sum

```

The summary table ranks the variables based on how different their average value is for potable and non-potable water. We use percent difference because the variables have different units and scales.

The following predictors have the largest average difference in potable/non-potable settings:

- Solids
- Chloramines
- Turbidity
- Organic Carbon
- pH
- Conductivity

# Visual Summaries
Given the number of predictors we have, it's probably best to investigate relationships through visual summaries. This will give us a better idea of the shape and spread of our data.

As a first step, we can produce pairwise plots using the `ggpairs()` function in the `ggally` package. We'll start with just examining the predictors we identified above:
```{r}
library(GGally)

# extract top predictors from numeric summary table
top_preds <- num_sum$name[1:5]

# drop na and only include top preds
water_top_preds <-
  water |>
  drop_na() |>
  select(potable, all_of(top_preds))

ggpairs(water_top_preds,
        mapping = aes(color = potable))
```

This matrix shows the relationships between each variable. The leftmost column shows histograms for each variable contingent on the potability of the water (non-potable = red, potable = blue). We'd like predictors with substantially different distributions for potable and non-potable water because they'd have better explanatory power for predicting potability.

Unfortunately, the histograms between potability and the other predictors share similar shapes. The top row box plots show a similar story, with the potable/non-potable box plots looking similar for all predictors.

Next, we examine interactions (the lower triangle off-diagonals). These scatter plots show us the separation of potable/non-potable water sources based on two predictors instead of one. Ideally we'd see two totally seperate groups of red/blue points, but there's some mingling going on.

The interactions with the greatest predictive power seem to be:

- `Solids` x `pH`
- `Solids` x `Chloramines`
- `Chloramines` x `Turbidity`

Again, `solids` seems to be a useful predictor of potability. We'll produce another pairwise plot to investigate the remaining predictors. I'm going to add `solids` and `chloramines` to the next one in case there are any relevant interactions:
```{r}
water_rest_preds <-
  water |>
  drop_na() |>
  select(potable, solids, chloramines, !any_of(top_preds))

ggpairs(water_rest_preds,
        mapping = aes(color = potable))
```

Checking out our cramped second matrix, we don't see any obvious stand-outs. It seems like the interactions with `solids` tend to do a decent job separating the potable from non-potable samples.

::: {.callout-note}
Note that the "separation" we see doesn't have to be a straight line through the predictor space, as both of the models we intend to fit are non-linear.
:::

I see the best separation in the following interactions:

- `solids` x `hardness`
- `solids` x `conductivity`
- `chloramines` x `conductivity`

# Recap of Useful Predictors

After examining the data, it seems like the following predictors will be most helpful for predicting potability:

- `solids`
- `chloramines`
- `conductivity`

Additionally, we saw some utility in the following set:

- `hardness`
- `ph`
- `turbidity`

With our exploratory analysis complete, we can start the modeling process!

[Click here for the Modeling Site](modeling.html)
