---
title: "ST 558 Final Project - Modeling"
author: Chris Grace
execute:
  warn: false
format: 
  html:
    toc: true
    theme: cosmo
---

[Click here for the EDA Site!](eda.html)

# Introduction

On the heels of the EDA we performed, we're ready to begin modeling!

Recall that our goal is to predict the potability (safeness to drink) of water based on some measurements about it. In the EDA, we found that the following predictors are most useful:

- `solids`
- `chloramines`
- `conductivity`
- `hardness`
- `ph`
- `turbidity`

## Preparing our Data

Before modeling, we'll prepare our data the same way as in the EDA file.

Additionally, we pare down to just the useful predictors.
```{r}
suppressPackageStartupMessages(library(tidyverse))

water_raw <- read.csv("water_potability.csv")

names(water_raw) <- tolower(names(water_raw))

# preds we identified in EDA step
useful_preds <- c("solids",
                  "chloramines",
                  "conductivity",
                  "hardness",
                  "ph",
                  "turbidity")

# prep dataset
water <-
  water_raw |>
  rename("potable" = "potability",
         "trihalo" = "trihalomethanes") |>
  mutate(potable = factor(potable,
                          labels = c("non_potable", "potable"))) |>
  select(potable, all_of(useful_preds))

# preview
head(water)
```
Looking good! This smaller data set will speed up computations.

We need to be careful though, we included `ph` even though it had some missing observations. 
```{r}
colSums(is.na(water))
```

There are `r sum(is.na(water$ph))` observations with missing `ph` readings. We didn't cover imputation, so it's most prudent to simply remove these observations from the data set:
```{r}
water <-
  water |>
  drop_na()

# no more missing ph values
sum(is.na(water$ph))
```

Now that the data is ready, we can partition our data into a train/test split!

## Train/Test Split + CV Folds

We're told to use a **70/30 train/test split**, so we reserve 70% of the data for model training. I'm including `potable` as the strata so that there's a proportional amount of potable/non-potable water in both partitions.

Also, we split up the training partition into **5 folds** as instructed.

```{r}
suppressPackageStartupMessages(library(tidymodels))

set.seed(558)

# 70/30 train/test split
water_split <- initial_split(water, prop = 0.70, strata = potable)

water_train <- training(water_split)
water_test  <- testing(water_split)

# 5 fold cross validation
water_folds <- vfold_cv(water_train, 5)
```

## Creating a Recipe

Next, we'll make a very simple recipe to add to our workflows.

This is a brief recipe for a few reasons:

- We don't have any dummy variables
- We don't need to normalize our numeric variables for tree-based methods
- We don't need to apply any transformations

Hence, we just specify that our response (potability) is a function of all the predictors in the data set.
```{r}
tree_rec <-
  recipe(potable ~ ., data = water_train)
```

At this point, it's time to work on each model separately!






























